---
title: 'Ch7: Likelihood and Max Likelihood'
author: "Jessica Couture"
date: "August 12, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## PC 7.1: Overexploitation of an unfished resource

"To describe this situation, we could use the Monte Carlo method to generate data in the time periods (starting with an unperturbed population), allow harvesting of half of the population at times 3, 4, and 5 (the overexploitation), and reduce the harvest rate to almlost zero for the last four time periods (the 'management action').  Assume: `r = 0.5, K = 1000, \sigma_{W} = 0.1, \sigma{V} = 0.1`"

```{r monteCarlo}

##### Define params:
r=0.5
K=1000
sigW=0.1 # process uncert
sigV=0.1 # observation uncert
ct=0.5
fshYrs<-c(3,4,5)


mcMod<-data.frame(t=0:10,
                  rlPop=c(K,rep(NA,10)))

logPOU<-function(df){
  
  df$Zw<-rnorm(11,mean=0,sd=1)
  for(i in 2:11){
    
    C=ifelse(df$t[i] %in% fshYrs,ct,0) # fishing just during the years defined
    sW<-exp(df$Zw[i]*sigW-(sigW^2)/2)
    
    df$rlPop[i]<-sW*(df$rlPop[i-1]+r*df$rlPop[i-1]*(1-(df$rlPop[i-1]/K))-(C*df$rlPop[i-1]))
    
  }
  
  df<-df %>%
    mutate(Zv=rnorm(11,mean=0,sd=1))%>%
    mutate(obsPop=rlPop*exp(Zv*sigV-(sigV^2)/2))
 return(df) 
}

modDF<-logPOU(mcMod)

plot(rlPop~t,data=modDF,type="l",col="green",xlab="year",ylab="pop",ylim=c(0,max(c(rlPop,obsPop))))
lines(obsPop~t,data=modDF, col="red")
```

## Log likelihood calculation: Only *observation* uncertainty exists

##### Calculate deterministic population abundances for varied r and K values

Build array to calculate deterministic populations for different r (2D matrices) and K (dimension 3 of array) over the 10yrs. 

```{r logLik-OU}

rs=seq(0.1,0.9,by=0.1)
Ks=seq(300,1500,by=100)

nllDF<-modDF%>%
  select(t,obsPop)

rsNA<-data.frame(Nr1=rep(NA,nrow(nllDF)),
                 Nr2=rep(NA,nrow(nllDF)),
                 Nr3=rep(NA,nrow(nllDF)),
                 Nr4=rep(NA,nrow(nllDF)),
                 Nr5=rep(NA,nrow(nllDF)),
                 Nr6=rep(NA,nrow(nllDF)),
                 Nr7=rep(NA,nrow(nllDF)),
                 Nr8=rep(NA,nrow(nllDF)),
                 Nr9=rep(NA,nrow(nllDF)))

nrDF<-cbind(nllDF,rsNA)

### logistic model function

logModLoop<-function(df,rw,ri,Ki){
       
  C=ifelse(df[rw,1] %in% fshYrs,ct,0) # fishing just during the years defined

  determ<-(df[rw-1,2]+ri*df[rw-1,2]*(1-(df[rw-1,2]/Ki))-(C*df[rw-1,2]))
  return(determ)
}

#########
determ<-function(rSeq,kSeq){
  
  detArr<-array(numeric(),c(nrow(nllDF),length(rs)+2,length(Ks)))
  
  for(j in 1:length(Ks)){
    
    inptMx<-as.matrix(nrDF)
    
    for(i in 1:length(rs)){

      # K<<-Ks[j]
      # r<<-rs[i]
      inptMx[2:nrow(nrDF),i+2]<-sapply(2:11,function(x) logModLoop(df=nrDF,rw=x,ri=rs[i],Ki=Ks[j]))

    }
    #print(inptMx)
    detArr[,,j]<-inptMx
  }
return(detArr)
}

detArr<-determ(rs,Ks)

```

##### Calculate deviation at each time period:
Eq. 7.36:
$$D_t = Z\sigma_{V}$$

```{r deviation}
devFun<-function(uncert){
  devArr<-array(numeric(),c(nrow(nllDF),length(rs)+2,length(Ks)))

  for(j in 1:length(Ks)){
    dev<-data.frame(t=0:10,
                    space=NA,
                    Dr1=NA,
                    Dr2=NA,
                    Dr3=NA,
                    Dr4=NA,
                    Dr5=NA,
                    Dr6=NA,
                    Dr7=NA,
                    Dr8=NA,
                    Dr9=NA)
    devMat<-as.matrix(dev)
    
  for(i in 1:length(rs)){
    devMat[2:nrow(nrDF),i+2]<-rnorm(10)*uncert
    
  }
      devArr[,,j]<-devMat
  }
return(devArr)
}

devArr<-devFun(sigV)
```

##### Calculate NLL at each time period:
Eq. 7.38:
$$L_t = log(\sigma_{V}) + \frac{1}{2}log(2\pi) + \frac{d_{t}^2}{2\sigma_{V}^2}$$

```{r deviation}
nllTFun<-function(uncert){
  nllTArr<-array(numeric(),c(10,length(rs),length(Ks)))

  for(j in 1:length(Ks)){
    ntd<-data.frame(Dr1=rep(NA,10),
                    Dr2=NA,
                    Dr3=NA,
                    Dr4=NA,
                    Dr5=NA,
                    Dr6=NA,
                    Dr7=NA,
                    Dr8=NA,
                    Dr9=NA)
    nllMat<-as.matrix(ntd)
    
  for(i in 1:length(rs)){
    nllMat[,i]<-sapply(1:10,function(x) log(uncert)+(1/2)*log(2*pi)+(devArr[x+1,i+2,j]^2/(2*uncert^2)))
    
  }
    #print(nllMat)
      nllTArr[,,j]<-nllMat
  }
return(nllTArr)
}

nllArr<-nllTFun(sigV)
```

##### Sum `nll_{t}` for all r and K combinations

```{r sumNLL}
nllDF2<-data.frame(r=rs,
                  k1=NA,
                  k2=NA,
                  k3=NA,
                  k4=NA,
                  k5=NA,
                  k6=NA,
                  k7=NA,
                  k8=NA,
                  k9=NA,
                  k10=NA,
                  k11=NA,
                  k12=NA,
                  k13=NA)
colnames(nllDF2)<-c("r",sapply(Ks,function(x) paste("K",x,sep="")))

for(i in 1:length(Ks)){
  nllDF2[,i+1]<-apply(nllArr[,,i],2, sum)
}

minDims<-which(nllDF2==min(nllDF2),arr.ind=T)

nll_r=nllDF2[minDims[1],1]
nll_k=colnames(nllDF2)[13]

plot(K1400~r,data=nllDF2,type="l",main="K=1400",ylab="NLL-obs uncert")
```


## Log likelihood calculation: Only *process* uncertainty exists

##### Calculate deterministic population abundances for varied r and K values

Build array to calculate deterministic populations for different r (2D matrices) and K (dimension 3 of array) over the 10yrs.  

```{r logLik-PU}

nllDF<-modDF%>%
  select(t,obsPop)

rsNA<-data.frame(Nr1=rep(NA,nrow(nllDF)),
                 Nr2=rep(NA,nrow(nllDF)),
                 Nr3=rep(NA,nrow(nllDF)),
                 Nr4=rep(NA,nrow(nllDF)),
                 Nr5=rep(NA,nrow(nllDF)),
                 Nr6=rep(NA,nrow(nllDF)),
                 Nr7=rep(NA,nrow(nllDF)),
                 Nr8=rep(NA,nrow(nllDF)),
                 Nr9=rep(NA,nrow(nllDF)))

nrDF<-cbind(nllDF,rsNA)

### logistic model function

logModLoopPU<-function(df,rw,ri,Ki){
  
  C=ifelse(df[rw,1] %in% fshYrs,ct,0) # fishing just during the years defined
  df$Zw<-rnorm(nrow(df))
  sW<-exp(df$Zw[rw]*sigW-(sigW^2)/2)

  determ<-sW*(df[rw-1,2]+ri*df[rw-1,2]*(1-(df[rw-1,2]/Ki))-(C*df[rw-1,2]))
  return(determ)
}

#########
determPU<-function(rSeq,kSeq){
  
  detArr<-array(numeric(),c(nrow(nllDF),length(rs)+2,length(Ks)))
  
  for(j in 1:length(Ks)){
    
    inptMx<-as.matrix(nrDF)
    
    for(i in 1:length(rs)){

      # K<<-Ks[j]
      # r<<-rs[i]
      inptMx[2:nrow(nrDF),i+2]<-sapply(2:11,function(x) logModLoopPU(df=nrDF,rw=x,ri=rs[i],Ki=Ks[j]))

    }
    #print(inptMx)
    detArr[,,j]<-inptMx
  }
return(detArr)
}

detArrPU<-determPU(rs,Ks)

```

##### Calculate deviation at each time period:
Eq. 7.40:
$$D_t = Z\sigma_{w}$$

```{r deviation-PU}

devArrPU<-devFun(sigW)
```

##### Calculate NLL at each time period:
Eq. 7.38:
$$L_t = log(\sigma_{W}) + \frac{1}{2}log(2\pi) + \frac{d_{t}^2}{2\sigma_{W}^2}$$

```{r nll-PU}


nllArrPU<-nllTFun(sigW)
```

##### Sum `nll_{t}` for all r and K combinations

```{r sumNLL-PU}
nllDF3<-data.frame(r=rs,
                  k1=NA,
                  k2=NA,
                  k3=NA,
                  k4=NA,
                  k5=NA,
                  k6=NA,
                  k7=NA,
                  k8=NA,
                  k9=NA,
                  k10=NA,
                  k11=NA,
                  k12=NA,
                  k13=NA)
colnames(nllDF3)<-c("r",sapply(Ks,function(x) paste("K",x,sep="")))

for(i in 1:length(Ks)){
  nllDF3[,i+1]<-apply(nllArrPU[,,i],2, sum)
}

minDims<-which(nllDF3==min(nllDF3),arr.ind=T)

nll_rPU=nllDF3[minDims[1],1]
nll_kPU=colnames(nllDF3)[13]

plot(nllDF3[,minDims[2]]~nllDF3$r,data=nllDF3,type="l",main=colnames(nllDF3)[minDims[2]],ylab="NLL-Proc. uncert",xlab="r")
```

## Monte Carlo: Generate data for eq. 7.43

"To illustrate the use of likelihood for model selstion, consier a model (Schnute 1987) relating the number of animals recorede by abservers in a survey (an index of abundance$I$)to the true abundance $D$ by

$$I = max\left(0,\frac{p+qD}{1+rD} \right)$$

Where $p$, $q$, and $r$ are parameters. We obtain a series of **nested** models by setting one or all of the parameters equal to 0. In the simplest case where $r =p = 0$, the index is proportional to the number of aminals present with constant of proportionality $q$, $I = qD$.  

The parameter $p$ allows for the possiblity that we may conclude that even when no animals are present some are recorded ($p>0$), or that we will not see any animals when they are rare ($p<0$). The parameter r allows for non-linearity between the index and the true abundance.  

Suppose that the number of individuals observed, $I_{obs}$, is the true number plus and observation uncertainty $V$ that is Poisson distributed. Thus, $I_{obs} = I+V$ will always equal or exceed the true number because $V\ge0$. As before, we begin by using Monte Carlo simulation to generate data in which we know the true situation:"

```{r montCarl7.4}

q=1.0
r=0.03
p=-3
D=1

schnt<-data.frame(D=1:20) %>%
  mutate(index=ifelse((p+q*D)/(1+r*D)>=0,(p+q*D)/(1+r*D),0)) %>%
  mutate(obs=sapply(.$index,function(x) x+rpois(1,4)))
  
trueMod<-lm(obs~D,schnt)

plot(obs~D,data=schnt)
lines(D~D,data=schnt)
abline(trueMod,lty=2)
```

## Likelihoods to compare models

Given a set of data generated by the previous pseudocode, we can estimate the likelihoods for each of the four models:

A. $I = max\left(0,qD \right)$
B. $I = max\left(0,\frac{p+qD}{1+D} \right)$
C. $I = max\left(0,\frac{qD}{1+rD} \right)$
D. $I = max\left(0,\frac{p+qD}{1+rD} \right)$

```{r likModSelect7.5}



```
