---
title: 'Chapter 8: Conservation Biology of the Wildbeest in  the Serengeti'
author: "Ignacia Rivera"
date: "January 23, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(magrittr)
library(here)
library(dplyr)
library(tidyr)
library(DescTools)
```


### The 1978 question
#### What happens with the wildbeest population once rainfall returns to normal (150 mm/year rather than 250 mm/year) ? 

##### *Competing hypotheses:*

1. The herd will collapse.
2. The herd will not collapse.

##### *Equations:*

1. Population growth - Logistic model

$$ N_{t+1} = N_t + rN_t ( 1 - \frac{N_t}{K})$$

2. We assume observation error

$$ N_{obs,t} = N_t + V_t$$ where $V_t$ is normally distirbuted with mean 0 and sd $\sigma_t$

3. Total food per ha ($T_t$)

$$T_t= 1.25 R_t $$
Where $R_t$ is the total dry season rainfall (in mm).


4. Food per animal ($F_t$)
 
 $$ F_t = \frac{T_t A}{N_t}$$
 
5. Births in a year ($B_t$)
 
 $$ B_t = 0.4 N_t$$
6. Survival Calf in a year ($s_{c, t}$)

$$ s_{c, t} = \frac{aF_t}{b+F_t} $$

Where $a \le 1$ is the maximum value of calf survival and $b$ is the value of food per individual at which survival is 50% of $a$. 

7. Survival Adult in a year ($s_{a, t}$)

$$ s_{a, t} = \frac{qF_t}{f+F_t} $$

Where $q \le 1$ is the maximum value of calf survival and $f$ is the value of food per individual at which survival is 50% of $a$. 

##### *Life histoy model:*

$$N_{t+1} = (s_{a, t})N_t + (s_{c, t})B_t $$
$$N_{t+1} = N_t (\frac{g 1.25 R_t/N_t}{f + 1.25R_t/N_t}) + 0.4 N_t(\frac{a 1.25R_t/N_t}{b + 1.25R_t/N_t})$$


At equilibirum: 

$$ N_{eq} = \frac{-b' + \sqrt{(b')^2 - 4a'c'}}{2a'}$$ where $a' = bf, b'= 1.25R (b+f-gb -0.4af)$, and $c' = (1.25)^2(1-g-0.4a)$

Adding harvesting after 1997:

$$  N_{t+1 = }\left\{
        \begin{array}{ll}
            (s_{a, t})N_t + (s_{c, t})B_t  & \quad t < 1977 \\
            (s_{a, t})N_t + (s_{c, t})B_t - h_t  & \quad t \geq 1977
        \end{array}
    \right.
$$

#### *Finding the best values for $r$ and $K$* 

Assuming the logistic model and observation uncertainty, and using the census data available in 1978.

```{r pseudocode 8.1}

# Initial parameters
R <- c(seq(from= 0.5, to=1, by= 0.25))
K <- c(seq(from= 300000, to = 3000000, by = 1000000))
N.0 <- 263000
years <- 1978- 1975

# Function that predicts based on logistic model and initial parameters

pred.pop.size <- function(R, K, N.0, years){

# Predicting Nt based on logistic model
  
## storing vector
pred.pop <- array(data= NA, dim=c(years, length(K), length(R)))
## setting starting year
pred.pop[1, , ] <- N.0

for (r in 1:length(R)) {
  
    for (y in 2:years)    {
    
      N.t.1 <- pred.pop[y-1, ,r] + R[r]*pred.pop[y-1, ,r] * (1-(pred.pop[y-1, ,r]/K)) #logistic growth

      pred.pop[y, , r] <- N.t.1

  }
  
} 

return(pred.pop)

}

pred.pop <- pred.pop.size(R, K, N.0, years)

# Calculating NLL for years with existing data for pop size estimate and standard deviation

# Data
census <- read.csv(here("ch_8/data/wildebeest_data.csv"))
colnames(census) <- c('year', 'rain', 'pop', 'sd', 'adult_mortality', 'calf_survival')

pop <- census %>% #dataframe with pop estimates and sigma per year
  filter(year < 1979 & pop != 'NA') %>% 
  mutate(sigma = ifelse(is.na(sd), 0.3, sd)) %>% 
  select(year, pop, sigma)

# Filttering years with estimates

pred.pop[,4 , ] <- years
pop <- pop %>%  
  left_join(pred.pop) %>% 
  mutate(nll = log(sigma) + 0.5* log(pi) + ((pop - pred.pop)^2)/(2*sigma^2))



```
