---
title: 'Eco-Detectives: chapter 11'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: FALSE
    toc_depth: 3
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/src/templates/ohara_hdr.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

source('https://raw.githubusercontent.com/oharac/src/master/R/common.R')

```

$\newcommand{\E}{\mathbb{E}} \newcommand{\Var}{\mathrm{Var}} \newcommand{\L}{\mathscr{L}} \newcommand{\LL}{\mathcal{L}}$

## Pseudocode 11.1: Understand the shape of the Goodness of Fit

1. Read in observed densities and index of abumdance (from chapter 7? where are these data?)
2. Determine lower and upper limits and step size for $q$
3. Determine lower and upper limits and step size for $p$
4. Set $p$ and $q$ at lower limit; set $r$ = 0.
5. Calc NLL of data given estimates of $p,q$.
6. (and 7.) Iterate over all values of $p, q$
8. Contour plot!

### Pseudocode 7.1 to generate data?

Monte Carlo simulation of unfished population that becomes overharvested, then unharvested again as a management action.

Assume parameters $r = 0.5, K = 1000, \sigma_W = 0.1, \sigma_V = 0.1$.

* Set initial pop at $K$.
* Use logistic eq'n with process uncertainty to determine pop size for subsequent years; in years 3, 4, 5, harvest 50% of population.
* Calc *observed* pop in each time period.
* Repeat for 10 years.

``` {r} 
r <- 0.5
K <- 1000
sigma_W <- 0.1
sigma_V <- 0.1

n_years <- 10
N_vec <- vector(length = n_years + 1)
N_vec[1] <- K

for(i in 1:n_years) { ### i <- 1
  N_t <- N_vec[i]
  ### calc process uncertainty for this step
  W_t <- exp(rnorm(1, mean = 0, sd = sigma_W) - sigma_W^2 / 2)
  
  ### calc harvest for this step: 50% in years 3, 4, 5; 0 else
  C_t <- ifelse(i %in% 3:5, 0.5 * N_t, 0)
  
  N_vec[i + 1] <- W_t * (N_t + r * N_t * (1 - N_t / K) - C_t)
  
}

### Since observed count is independent of process uncertainty, we can just
### apply the observation uncertainty after the fact more easily.
V_vec <- exp(rnorm(length(N_vec), mean = 0, sd = sigma_V) - sigma_V^2 / 2)
N_obs <- N_vec * V_vec

df <- data.frame(year = 1:(n_years + 1),
                 actual   = N_vec,
                 observed = N_obs) %>%
  gather(type, N, -year)
ggplot(df, aes(x = year, y = N, color = type)) +
  ggtheme_plot() +
  xlim(c(0, NA)) + ylim(c(0, NA)) +
  geom_point() +
  geom_line()

```


