---
title: 'Eco-Detectives: chapter 3'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/src/templates/ohara_hdr.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

source('https://raw.githubusercontent.com/oharac/src/master/R/common.R')

```

# Chapter 3: Probability and Probability Models

## Pseudocode 3.1

Here we're implementing a binomial probability distribution to plot the probability of finding $k$ successes out of $N$ attempts.  Eqn 3.47 gives the probability of zero successes:

$$p(0, N) = (1 - p)^N$$
Eqn 3.48 gives the recursive formula to find $k$ successes based on the probability of finding $k - 1$ successes:
\begin{align*}
  p(k, N) &= \binom{N}{k} p^k (1 - p)^{N - k}\\
    &= \left(\frac{N - k + 1}{k}\right) \left(\frac{p}{1-p}\right) p(k-1, N)
\end{align*}

Start by choosing values for $p$ and $N$.  Let's start with $p = 0.25$ and $N = 10$.

``` {r ch 3 pseudocode 3-1}

p <- 0.25
N <- 10

k_vec <- 0:N
prob <- rep(NA, length(k_vec))

for (i in seq_along(k_vec)) {
  k <- k_vec[i]
  prob[i] <- ifelse(k == 0,
                    (1 - p)^N, 
                      ### k = 0, so use 3.47
                    (N - k + 1)/(k) * (p)/(1 - p) * prob[i-1])
                      ### k > 0, so use 3.48 and refer to previous p(k-1, N)
}
binomial_df <- data.frame(k = k_vec, p = prob)
                      
ggplot(binomial_df, aes(x = k, y = p)) +
  ggtheme_plot() +
  geom_line() +
  geom_point()
```

Let's try non-iteratively...

``` {r ch 3 pseudocode 3-1 vectorized}

N <- 100
p_vals <- c(0.1, 0.25, 0.5, 0.75, .97)

bin2_df <- data.frame(k = rep(0:N, times = length(p_vals)),
                     p = rep(p_vals, each = N + 1)) %>%
  mutate(prob = ifelse(k == 0,
                      (1 - p)^N, 
                        ### k = 0, so use 3.47
                      choose(N, k) * p^k * (1 - p)^(N - k)))
                        ### k > 0, so use the original eq 3.44
                      
ggplot(bin2_df, aes(x = k, y = prob, color = p)) +
  ggtheme_plot() +
  geom_line(aes(group = p)) +
  geom_point()
```

Note there's a function `dbinom()` (and `pbinom`, `qbinom`, `rbinom`) that generate binomial density, distribution, quantile, and random numbers for a binomial distribution with `x`/`q`/`p`/`n` of outputs plus `size` (# trials) and `prob` (probability of success on each trial).

## Pseudocode for multinomial

Instead of 3.44 use eq. 3.50 in the code; if we want it as a data frame, instead of using several values for $p$ across a single $k_1$, we can use different values for $k_2, k_3, ...$.  Probably makes more sense to do it as a matrix if there are more than two or three $k_x$ values.

## Pseudocode 3.2

Here we're dealing with the Poisson distribution

$$p(k, t) = \frac{\lambda^k e^{-\lambda}}{k!} = \frac{(rt)^k e^{-rt}}{k!}$$

Specify $r$, $t$, and $cutoff$.  Since $rt$ takes the place of $\lambda$, $\mathbb{E}[x] = rt$ and $Var(x) = rt$.  We'll iterate over values of $k$ until the cumulative probability gets close to 1, i.e. the CDF instead of the PDF.

```{r pseudocode 3-2}

r <- .05
t <- 40
cutoff <- 0.9999
k <- NA; prob = NA

i <- 1
k[1] <- 0
prob[1]  <- exp(-r*t)
sum_p <- prob[1]

while(sum_p < cutoff) {
  i <- i + 1
  k[i] <- k[i-1] + 1
  prob[i] <- r * t / k[i] * prob[i - 1]
  
  sum_p <- sum_p + prob[i]
}

poisson_df <- data.frame(k = k, prob = prob)

ggplot(poisson_df, aes(x = k, y = prob)) + 
  ggtheme_plot() +
  geom_point() +
  geom_line()
```

try non-iteratively

```{r pseudocode 3-2 vectorized}

r_vec <- c(0.01, 0.02, 0.05, .10, .25, .5)
t <- 40
cutoff <- 0.9999
k_max <- 100

poisson_df <- data.frame(k = rep(0:k_max, times = length(r_vec)), 
                         r = rep(r_vec, each = k_max + 1),
                         prob = NA) %>%
  mutate(prob = exp(-r * t) * (r * t)^k / factorial(k)) %>%
  group_by(r) %>%
  mutate(CDF = cumsum(prob)) %>%
  ungroup() %>%
  filter(CDF < cutoff)

ggplot(poisson_df, aes(x = k, y = prob, color = r)) + 
  ggtheme_plot() +
  geom_point() +
  geom_line(aes(group = r))
```
