---
title: "Chapter 3"
author: "Molly Wilson"
date: "6/22/2018"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    number_sections: false
---

```{r, echo=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
```

# Chapter 3: Probability and Probability Models

## 3.1 Binomial distribution

In this excercise we are plotting the probability distribution of getting $k$ successes out of $N$ attempts based on a given probability $p$ of getting a success on any given attempt.

1. Specify $p$ and $N$  
$p$ = 0.1  
$N$ = 7  

2. Find $p(0,N)$ from equation 3.47
$$p(0, N) = (1 - p)^N$$
$$p(0,N) = (1-0.1)^7 = (0.9)^7 = 0.47$$

3. Find $p(k,N)$ from equation 3.48  
$$p(k.n)=\frac{N!}{(k!*(N-k!))}*p^k*(1-p)^{N-k}$$

For-loop method (with hints from Casey's code):
```{r}
p <- 0.1
N <- 7

k_vec <- 0:N # create vector of k values
prob <- rep(NA, length(k_vec)) # create (currently empty) vector to store probability values

for (i in seq_along(k_vec)) {
  k <- k_vec[i]
  prob[i] <- ifelse(k == 0,
                    (1 - p)^N, # using equation 3.47 for when k=0
                    (N - k + 1)/(k) * (p)/(1 - p) * prob[i-1]) # when k>1, p(k,N) is based of p(k-1,N) using equation 3.48
}
binomial_df <- data.frame(k = k_vec, p = prob)
                      
ggplot(binomial_df, aes(x = k, y = p)) +
  geom_line() +
  geom_point()
```

Vectorized version (also help from Casey):
``` {r ch 3 pseudocode 3-1 vectorized}
N <- 7
p_vals <- c(0.1, 0.25, 0.5, 0.75, .97)

binomial_df2 <- data.frame(k = rep(0:N, times = length(p_vals)),
                     p = rep(p_vals, each = N + 1)) %>%
  mutate(prob = ifelse(k == 0,
                      (1 - p)^N, 
                      choose(N, k) * p^k * (1 - p)^(N - k))) # using equation 3.44 instead of 3.47 because we are not basing p(k,N) on p(k-1,N)
                      
ggplot(binomial_df2, aes(x = k, y = prob, color = p)) +
  geom_line(aes(group = p)) +
  geom_point()
```

Cheating version using dbinom
```{r}
k <- 1:7
k <- as.data.frame(k)

binomial_df3 <- as.data.frame(apply(k, 2, dbinom, size=7, prob=0.1)) %>% select(p=k) %>% cbind(k) %>% select(k,p)
# still need to bind row with p(0,k)

ggplot(binomial_df2, aes(x = k, y = p)) +
  geom_line() +
  geom_point()
```

## 3.2 Poisson distribution

In this excercise we are using the Poisson distribution, which is essentially the binomial distribution when the time period $t$ gets infinitely small, and using a rate value $r$ instead of $p$ that defines the rate of successes.

1. Choose $r$, $t$, and $cutoff$  
$r$ = 0.1  
$t$ = 20  
$cutoff$ = 0.999  

2. Set: 
$$p(0,t)=e^{-rt}$$
$$SUM=p(0,t)$$

3. Find $p(k,t)$ from equation 3.57
$$p(k, t) = \frac{\lambda^k e^{-\lambda}}{k!} = \frac{(rt)^k e^{-rt}}{k!}$$

While-loop version (adapted from Casey's code):
```{r}
r <- .1
t <- 20
cutoff <- 0.9999
k <- NA; prob = NA

# set starting conditions for the first iteration:
i <- 1
k[1] <- 0
prob[1]  <- exp(-r*t)
sum_p <- prob[1]

while(sum_p < cutoff) {
  i <- i + 1 # increase i with each step
  k[i] <- k[i-1] + 1 # increase k with each step
  prob[i] <- r * t / k[i] * prob[i - 1] # equation 3.57
  
  sum_p <- sum_p + prob[i] # increase value of "sum_p" with each iteration to get cumulative probability value
}

poisson_df <- data.frame(k = k, prob = prob)

ggplot(poisson_df, aes(x = k, y = prob)) + 
  geom_point() +
  geom_line()
```

Vectorized version (from Casey):
```{r}
r_vec <- c(0.01, 0.02, 0.05, .10, .25, .5)
t <- 20
cutoff <- 0.9999
k_max <- 100

poisson_df2 <- data.frame(k = rep(0:k_max, times = length(r_vec)), # creating repeating k values for each given r value
                         r = rep(r_vec, each = k_max + 1)) %>% # don't understand each vs. times...?
                  
  mutate(prob = exp(-r * t) * (r * t)^k / factorial(k)) %>% # calculating p(k,t) without knowing p(k-1,t)
  group_by(r) %>%
  mutate(CDF = cumsum(prob)) %>%
  ungroup() %>%
  filter(CDF < cutoff)

ggplot(poisson_df2, aes(x = k, y = prob, color = r)) + 
  geom_point() +
  geom_line(aes(group = r))
```


## 3.3 Monte Carlo: random variable from a binomial distribution

The Monte Carlo method uses random number generators to construct data. In this excercise we're using a uniform distribution to select an integer result from a binomial distribution.

1. Specify parameters $N$ and $p$, and choose uniformly distributed random number U

$N$=15
$p$=0.1
$k$=0
SUM=0

2. Compute $p(k,N)$ from equation 3.44:

$$p(k,N)=choose(N, k) * p^k * (1 - p)^(N - k)$$

Dataframe/lapply method (adapted from Casey):
```{r}
N <- 15
p_vec <- c(.1,.2,.5,.9)
U<-runif(15)

randbinom <- function(N, k, p, U) {
  randbinom_df <- data.frame(N = N,
                          k = rep(0:N, times = length(p_vec)),
                          p = rep(p_vec, each = N + 1)) %>%
    group_by(p) %>%
    mutate(pdf = choose(N, k) * p^k * (1 - p)^(N - k),
           cdf = cumsum(pdf)) %>%
    filter(cdf > U) %>%
    summarize(U = U, N = first(N), k_successes = first(k))

  return(randbinom_df)
}

randbinom_df <- lapply(U, FUN = function(x) {
  randbinom2(N, k, p_vec, x)
}) %>%
  bind_rows()

ggplot(randbinom_df, aes(y = U, x = k_successes, color = p)) +
  geom_point()
```


## 3.4 An ecological scenario: the simple population model with process and observation uncertainty

$$N_{t+1} = sN_t + b_t + \sigma_w$$ 
$$N_{obs,t} = N_t + \sigma_v$$ 

1. Specify $s$,$b$,$\sigma_N$, $\sigma_w$,$\sigma_v$,and $N_0$
$s$= survival = 0.8
$b$= births = 20
$N_0$= starting population = 50
$\sigma_N$= 
$\sigma_w$= process uncertainty = 10
$\sigma_v$= observation uncertainty = 0

2. Begin a loop over 50 time steps

```{r}
pop <- function (s,b,N0,sigma_N,sigma_w,sigma_v) {
  t_vec <- 1:50
  N_act <- rep(NA, 50)
  N_obs <- rep(NA, 50)

for(t in t_vec){
  if(t==1){ # set actual and observed populations in initial time step to N0
    N_act<-N0
    N_obs<-N0
  }
  else{
  N_act[t]<-s*N_act[t-1]+b+rnorm(1, 0, sigma_w)
  N_obs[t]<-N_act[t]+rnorm(1, 0, sigma_v)
  }
}
  N_df <- data.frame(t = t_vec,
                     N_act = N_act,
                     N_obs = N_obs)
  return(N_df)
}

```

Generating values with no observation uncertainty ($\sigma_v$=0)
```{r}
pop1<-pop(s= 0.8, b = 20, sigma_N=0, sigma_w=10, sigma_v=0, N0=50) %>% mutate(N_act_prev = lead(N_act), N_obs_prev=lead(N_obs)) # create values for N_act and N_obs in previous time step for graphing purposes

ggplot(pop1, aes(y = N_act, x = N_act_prev)) + geom_smooth(method=lm) + geom_point()
```

Generating values with observation uncertainty ($\sigma_v$=10)
```{r}
pop2<-pop(s= 0.8, b = 20, sigma_N=0, sigma_w=10, sigma_v=10, N0=50) %>% mutate(N_act_prev = lead(N_act), N_obs_prev=lead(N_obs))

ggplot(pop2, aes(y = N_obs, x = N_obs_prev)) + geom_smooth(method=lm) + geom_point()
```


