   ---
title: 'Eco Detectives: Ch4 work'
author: "Jessica Couture"
date: "July 2, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# A Monte Carlo approach for estimating the change of success in an observer program

We can understand the potential for failure caused bt ignoring aggregation by asking how likely one is to obtain statistically meaningful data for a given level of observer coverage. This question can be answered using a Monte Carlo method...

$$Pr(C_{i}=c)=p(c)=\frac{\Gamma(k+c)}{\Gamma(k)c!}\left( \frac{k}{k+m} \right)^{k}\left( \frac{m}{m+k} \right)^{c}$$

```{r pseudCode4.1}

byctch<-read.csv("data/table_4_3.csv")%>%
  rename(nTows=frequency,nBrds=bycatch) %>%
  mutate(totBrds=nBrds*nTows) %>%
  #mutate(propTow=(bycatch*frequency/sum(frequency)))
  summarise(m=sum(totBrds)/sum(nTows),
            s=sum(nTows*(nBrds-m)^2)/(sum(nTows)-1))

## calculate mean
m=sum(byctch$totBrds)/sum(byctch$frequency) # weighted means 
#m2=sum(byctch$propTow)/sum(byctch$frequency) # this is waaaay too small

Ntow=1000
Nsim=150
mDat=byctch$m
kDat=(mDat^2)/(byctch$s-mDat)
d=0.25
tq=qt(0.95,Ntow-1) # qt(1-alpha/2,df=N-1

```

Calculate probabilities of $C_{i}$ using the $\Gamma$ distribution

```{r pseudCode4.2}

prC<-function(c,m=mDat,k=kDat){
  p=(gamma(k+c)/(gamma(k)*factorial(c)))*(k/(k+m))^k*(m/(m+k))^c
  return(p)
}

sim1=data.frame(bc=c(0:50),
                probBc=sapply(0:50,function(x) prC(x))) %>%
  mutate(randUnif=runif(51,0,1))%>%
  mutate(cuSum=cumsum(probBc))

plot(sim1$bc,sim1$cuSum,type="l",
     xlab="bycatch level",ylab="cumulative prob")

```

Apply random numbers to the distribution function

```{r loopSims}
simTow<-matrix(data=NA,nrow=Ntow,ncol=Nsim)

for(i in 1:Ntow){
  rn<-runif(1,0,1)
  
}
```

### Here I am stuck, can we walk through this?
