   ---
title: 'Eco Detectives: Ch4 work'
author: "Jessica Couture"
date: "July 2, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# A Monte Carlo approach for estimating the change of success in an observer program

We can understand the potential for failure caused bt ignoring aggregation by asking how likely one is to obtain statistically meaningful data for a given level of observer coverage. This question can be answered using a Monte Carlo method...

$$Pr(C_{i}=c)=p(c)=\frac{\Gamma(k+c)}{\Gamma(k)c!}\left( \frac{k}{k+m} \right)^{k}\left( \frac{m}{m+k} \right)^{c}$$

```{r pseudCode4.1}

byctch<-read.csv("data/table_4_3.csv")%>%
  rename(nTows=frequency,nBrds=bycatch) %>%
  mutate(totBrds=nBrds*nTows) %>%
  #mutate(propTow=(bycatch*frequency/sum(frequency)))
  summarise(m=sum(totBrds)/sum(nTows),
            s=sum(nTows*(nBrds-m)^2)/(sum(nTows)-1))

## calculate mean
m=sum(byctch$totBrds)/sum(byctch$frequency) # weighted means 
#m2=sum(byctch$propTow)/sum(byctch$frequency) # this is waaaay too small

Ntow=5000
Nsim=150
mDat=byctch$m
kDat=(mDat^2)/(byctch$s-mDat)
d=0.25
tq=qt(0.95,Ntow-1) # qt(1-alpha/2,df=N-1

```

Calculate probabilities of $C_{i}$ using the $\Gamma$ distribution

```{r pseudCode4.2}

prC<-function(c,m=mDat,k=kDat){
  p=(gamma(k+c)/(gamma(k)*factorial(c)))*(k/(k+m))^k*(m/(m+k))^c
  return(p)
}

sim1=data.frame(bc=c(0:50),
                probBc=sapply(0:50,function(x) prC(x))) %>%
  #mutate(randUnif=runif(51,0,1))%>% # make this in the loop
  mutate(cuSum=cumsum(probBc)) %>%
  mutate(cuSumLo=c(0,cuSum[1:nrow(.)-1])) ## add lagged column to make the intervals (also could use "lag" function)

plot(sim1$bc,sim1$cuSum,type="l",
     xlab="bycatch level",ylab="cumulative prob")

```

##### Try one simulation of tows
Simulate how many birds would be caught for each randomly produced tow based on our data distribution

```{r oneSim}
simVec<-vector(length=Ntow) # make this a column in a vector below
rn<-runif(n=Ntow,0,1) # run this in the simulation loop below to get a new set of random #s for each simulation
  
for(i in 1:Ntow){
  simVec[i]<-which(rn[i]>sim1$cuSumLo & rn[i]<sim1$cuSum)-1
} #

m1=sum(simVec)/Ntow # or mean(simVec)

s1=sum(sapply(simVec,function(x) x-m1)^2)/(Ntow-1) #or sd(simVec)^2

```

##### Loop through additional simulations (Nsim)

```{r loopSims}
simTow<-matrix(data=NA,nrow=Ntow,ncol=Nsim)

mnsVec<-vector(length = Nsim)
varVec<-vector(length = Nsim)
rngVec<-vector(length = Nsim)

for(j in 1:Nsim){
  rn<-runif(n=Ntow,0,1)
  
  for(i in 1:Ntow){
    simTow[i,j]<-which(rn[i]>sim1$cuSumLo & rn[i]<sim1$cuSum)-1
  
  }
  
  mnsVec[j]<-sum(simTow[,j])/Ntow
  varVec[j]<-sum(sapply(simTow[,j],function(x) x-mnsVec[j])^2)/(Ntow-1) 
  #rngVec[j]<-2*sqrt(varVec[j]/Ntow)*tq
  
}
```

## Probability of success

Successes occur when the range smaller the defined uncertainty level: d x mean. 

```{r probSucc}

prSuc<-data.frame(means=mnsVec,
                  vaR=varVec)%>%
  mutate(rngs=2*sqrt(vaR/Ntow)*tq,
         criteria=means*d,
         succ=rngs<criteria)

sum(prSuc$succ)
```

### Plot the probabilties of success for different numbers of tows

Will need to loop through the above simulations and statistics for a variety of different Ntow values. So adding an additional outside loop to loop throught the different values. 

```{r probSuccPlt}
ntowVec<-c(0:5000)

simArry<-array(data=NA,dim=c(Ntow,Nsim,length(ntowVec)))

summArry<-matrix(data=NA,nrow=Nsim,ncol=length(ntowVec))

for(k in 1:length(ntowVec)){
  
  for(j in 1:Nsim){
    rn<-runif(n=Ntow,0,1)
  
    for(i in 1:Ntow){
      simTow[i,j]<-which(rn[i]>sim1$cuSumLo & rn[i]<sim1$cuSum)-1
  }
  
  summArry[j,k]<-sum(simTow[,j])/ntowVec[k]
  varVec[j]<-sum(sapply(simTow[,j],function(x) x-mnsVec[j])^2)/(Ntow-1) 
  #rngVec[j]<-2*sqrt(varVec[j]/Ntow)*tq
  
}

prSuc<-data.frame(means=mnsVec,
                  vaR=varVec)%>%
  mutate(rngs=2*sqrt(vaR/Ntow)*tq,
         criteria=means*d,
         succ=rngs<criteria)

sum(prSuc$succ)
}

```

...NOT complete, look at this and replace old code with stuff that works
