---
title: 'Eco-Detectives: chapter 6'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/src/templates/ohara_hdr.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

source('https://raw.githubusercontent.com/oharac/src/master/R/common.R')

```

$\newcommand{\E}{\mathbb{E}}\newcommand{\Var}{\mathrm{Var}}$

# Chapter 6: Insect oviposition behavior

## 6.1

$$\text{SSQ}(c_f) = \sum_{E=1}^{23} \sum_{C=1}^4 (C - c_f)^2 N(E, C)$$

Read data into table.  The pseudocode looks like three nested loops: 

* cycle over $c_f \in [1, 4]$
* cycle over $E \in [4, 23]$
* cycle over $C \in [1, 4]$
* replace $SSQ(c_f)$ by $SSQ(c_f) + (C - c_f)^2 N(E, C)$.  Then cycle over $E$ and $C$ for thi $c_f$.

``` {r} 

N <- read_csv('table_6_1.csv') %>%
  clean_df_names()

N_c <- N[ , 2:5] %>% sum()
  ### sum up all the observations
  
SSQ <- rep(0, 4)
for(c_f in 1:4) { ### c_f <- 1
  for(E in 4:23) { ### E <- 12
    for(C in 1:4) { ### C <- 3
      N_E_C <- N[N$eggcomplement == E, C + 1] %>% as.integer()
        ### C matches the columns, skipping the eggcomplement column
      SSQ[c_f] <- SSQ[c_f] + (C - c_f)^2 * N_E_C
    }
  }
}

SSQ_N <- SSQ / N_c

print(SSQ_N)
# 2.6862745 0.6274510 0.5686275 2.5098039
```


``` {r} 

N <- read_csv('table_6_1.csv') %>%
  clean_df_names()

N_c <- N[ , 2:5] %>% sum()
  ### sum up all the observations
  
N_df <- N %>%
  gather(C, n_obs, starts_with('n')) %>%
  mutate(C = str_replace_all(C, '[^0-9]', '') %>% as.integer(),
         c_f = list(c(1, 2, 3, 4))) %>%
  unnest(c_f)

SSQ_df <- N_df %>%
  group_by(c_f) %>%
  summarize(ssq = sum((C - c_f)^2 * n_obs)) %>%
  mutate(ssq_n = ssq / N_c)

knitr::kable(SSQ_df)
```

## 6.2

$$\text{SSQ (single switch)} = \sum_{E=1}^{23} \sum_{C=1}^4 [C - c(E)]^2 N(E, C)$$
where

\begin{align*}
  c(e) = \begin{cases} 
    c_1 &\text{if } e \leq e_1\\
    c_2 &\text{if } e > e_1
  \end{cases}
\end{align*}

Let's stick with the loop-centric version...

``` {r} 

N <- read_csv('table_6_1.csv') %>%
  clean_df_names()

N_c <- N[ , 2:5] %>% sum()
  ### sum up all the observations
  
SSQ_opt <- 1e100
  ### initialize to an absurdly high sum of squares
c1_opt <- c2_opt <- e1_opt <- 0
  ### don't really need to initialize these...

for(c1 in 1:4) { ### c1 <- 1
  for(c2 in c1:4) { ### c2 <- 2 ### assume c2 is greater than c1 for larger egg complement
    for(e1 in 1:23) {
      # cat('c1 = ', c1, '; c2 = ', c2, '; e1 = ', e1, '\n')
      SSQ <- 0
        ### initialize SSQ for this parameter set
      
      for(E in 4:23) { ### E <- 12
        for(C in 1:4) { ### C <- 3
          N_E_C <- N[N$eggcomplement == E, C + 1] %>% as.integer()
            ### C matches the columns, skipping the eggcomplement column
          
          c_e <- ifelse(E <= e1, c1, c2)
          
          SSQ <- SSQ + (C - c_e)^2 * N_E_C
        } ### end of E loop
      } ### end of C loop - i.e. end of SSQ calcs for these e1, c1, c2 vals
      
      ### Check SSQ vs SSQ_opt - if better (lower), set c1_opt, c2_opt, e1_opt
      ### to current values.
      if(SSQ < SSQ_opt) {
        SSQ_opt <- SSQ
        c1_opt  <- c1
        c2_opt  <- c2
        e1_opt  <- e1
      }
    } ### end of e1 loop
  } ### end of c2 loop
} ### end of c1 loop

SSQ_N <- SSQ_opt / (N_c - 6)

cat(sprintf('Optimal values: c1 = %d, c2 = %d, e1 = %d, SSQ = %f\n', 
            c1_opt, c2_opt, e1_opt, SSQ_N))
# 2.6862745 0.6274510 0.5686275 2.5098039
```

### Likelihood profile for switching egg complement

Holding $c_1$ and $c_2$ at their optimal values, vary $e_1$ to generate a likelihood profile.

``` {r} 

N <- read_csv('table_6_1.csv') %>%
  clean_df_names()

N_c <- N[ , 2:5] %>% sum()
  ### sum up all the observations

c1_opt <- 2
c2_opt <- 3
  
N_df <- N %>%
  gather(C, n_obs, starts_with('n')) %>%
  mutate(C = str_replace_all(C, '[^0-9]', '') %>% as.integer(),
         e1 = list(c(4:23))) %>%
  unnest(e1) %>%
  mutate(c_e = ifelse(eggcomplement <= e1, c1_opt, c2_opt))

SSQ_df <- N_df %>%
  group_by(e1) %>%
  summarize(ssq = sum((C - c_e)^2 * n_obs)) %>%
  mutate(ssq_n = ssq / (N_c - 6))

ggplot(SSQ_df, aes(x = e1, y = ssq_n)) +
  ggtheme_plot() +
  geom_point() +
  xlim(c(0, NA)) +
  ylim(c(0, NA)) +
  geom_vline(xintercept = 8, color = 'red')
```

## 6.3 bootstrapping

``` {r} 

N <- read_csv('table_6_1.csv') %>%
  clean_df_names()

N_c <- N[ , 2:5] %>% sum()
  ### sum up all the observations

c1_opt <- 2
c2_opt <- 3
  
### let's make it truly long format
N_df <- N %>%
  gather(C, n_obs, starts_with('n')) %>%
  mutate(C = str_replace_all(C, '[^0-9]', '') %>% as.integer()) %>%
  tidyr::uncount(n_obs)

### initialize counts for fixed model comparison (fixed = 2 vs 3)
### and switch model comparison (vs both fixed)
n_iterations <- 10000
fixed <- switch <- rep(NA, n_iterations)
  
ptm <- proc.time()

for (i in 1:n_iterations) { ### i <- 1
  if(i %% 100 == 0) message(i, '... ')
  N_bs <- N_df %>%
    sample_frac(size = 1, replace = TRUE) %>%
    count(eggcomplement, C)
  
  SSQ_c2 <- N_bs %>%
    mutate(c_f = 2) %>%
    summarize(ssq = sum((C - c_f)^2 * n)) %>%
    mutate(ssq_n = ssq / (N_c)) %>%
    .$ssq_n
  
  SSQ_c3 <- N_bs %>%
    mutate(c_f = 3) %>%
    summarize(ssq = sum((C - c_f)^2 * n)) %>%
    mutate(ssq_n = ssq / (N_c)) %>%
    .$ssq_n
  
  SSQ_switch <- N_bs %>%
    mutate(c_e = ifelse(eggcomplement <= 8, 2, 3)) %>%
    summarize(ssq = sum((C - c_e)^2 * n)) %>%
    mutate(ssq_n = ssq / (N_c)) %>%
    .$ssq_n
  
  fixed[i]  <- ifelse(SSQ_c2 < SSQ_c3, 'c_f = 2', 'c_f = 3')
  switch[i] <- case_when(SSQ_switch < (min(SSQ_c2, SSQ_c3)) ~ 'switch',
                         SSQ_c2 < SSQ_c3                    ~ 'c_f = 2',
                         TRUE                               ~ 'c_f = 3')
}

cat(sprintf('%s iterations: elapsed time: %.2f \n', n_iterations, (proc.time() - ptm)[3]))
table(fixed)
table(switch)
```


