---
title: 'Ch6: model selection with SSQ'
author: "Jessica Couture"
date: "August 9, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Egg independent clutches

##### Calculate weighted sums of squares

"Weight the sum of squared deviations by the number of observations at that egg complement (E) and clutch (C). There are 4 values of clutch (`c_{f} = 1, 2, 3, or 4`), each of which prodices a different value of SSQ(`c_{f}`)."


```{r 6.1data}
ecDat<-read.csv("table_6_1.csv")

head(ecDat)

```

##### Functions to loop through SSQ calcs

$$SSQ(c_{f}) = \sum_{E=1}^{23}\sum_{C=1}^{4}(C-c_f)^{2}N(E,C) $$

For each `c_{f}` first loop through `C` values to calculate sums and then loop through E values. The pseudocode says to use `E=4-23`.

```{r ssqLoops}

wtSSQ<-function(cf){
 
  sum2<-0
  
  for(i in 1:nrow(ecDat)){
  
    sum1<-0
      
    for(j in 1:4){
        eq<-((j-cf)^2)*(ecDat[i,j+1])
        sum1<-sum1+eq
    }
  
    sum2<-sum2+sum1
  }
  return(sum2)
}

ssq1=wtSSQ(1)
ssq2=wtSSQ(2)
ssq3=wtSSQ(3)
ssq4=wtSSQ(4)

ssqDF<-data.frame(cf=c(1,2,3,4),
                  ssq=c(ssq1,ssq2,ssq3,ssq4)) %>%
  mutate(compare=ssq/sum(ecDat[,2:ncol(ecDat)]))

ssqDF

```

## Clutch size increases with complement size

"We denote clutch size when egg complement is `e` by `c(e)`. Adopting this viewpoint, the simplest variable-vlutch model is one in which the parasitoid switches  from clutch size `c_{1}` to clutch size `c_2 > c_1` at egg complement `e_1`"  

In this case the SSQ is calculated with:
$$SSQ(switch) = \sum_{E=1}^{23}\sum_{C=1}^{4}(C-c(E))^{2}N(E,C) $$

Where `c(E)` takes on either `c_1` or `c_2` depending on the values of `E` and `e_1`

##### Calculate sum of squares

```{r ssqSwtch}

c1<-1:4

swiDf<-data.frame(matrix(nrow=20,ncol=4))

swiSSQ<-function(clch,swi){
 
  c1<-clch
  c2<-clch+1
  
  sum2<-0
  
  for(i in 1:nrow(ecDat)){
  
    sum1<-0
      
    for(j in 1:4){
      ce<-ifelse((i+3)>swi,c2,c1)
      
      eq<-((j-ce)^2)*(ecDat[i,j+1])
      
      sum1<-sum1+eq
    }
  
    sum2<-sum2+sum1
  }
  return(sum2)
}

swiDf[,1]<-sapply(ecDat$EggComplement,function(x) swiSSQ(1,x))
swiDf[,2]<-sapply(ecDat$EggComplement,function(x) swiSSQ(2,x))
swiDf[,3]<-sapply(ecDat$EggComplement,function(x) swiSSQ(3,x))
swiDf[,4]<-sapply(ecDat$EggComplement,function(x) swiSSQ(4,x))

colnames(swiDf)=c("C1is1","C1is2","C1is3","C1is4")

plot(ecDat$EggComplement,swiDf$C1is1,type="l")
plot(ecDat$EggComplement,swiDf$C1is2,type="l",col="red")
plot(ecDat$EggComplement,swiDf$C1is3,type="l",col="blue")
plot(ecDat$EggComplement,swiDf$C1is4,type="l",col="green")

which.min(swiDf$C1is2)+3
```

## Bootstrapping

```{r transfData}
rawDat<-data.frame(E=numeric(), C=numeric())

for(i in 1:nrow(ecDat)){
  for(j in 2:ncol(ecDat)) {
    newDF<-data.frame(E=rep(ecDat[i,1],ecDat[i,j]),
                      C=rep(j-1,ecDat[i,j]))
    rawDat<-rbind(rawDat,newDF)
  }
}
head(rawDat)
```

Use `sample()` to draw the bootstrap samples from the `rawDat` dataframe:

```{r sample}

rawVec<-apply(rawDat,1,function(x) paste(x[1],x[2]))

bsVec<-sample(rawVec,replace = T)

bsDf<-data.frame(E=sapply(strsplit(bsVec," "),function(x) x[1]),
                 C=sapply(strsplit(bsVec," "),function(x) x[2]))
```

##### Test bootstrapped data using the tests above:

Fixed clutch model:

```{r fixdClutch}
ecDat<-data.frame(table(bsDf)) %>%
  spread(key=C,value=Freq) %>%
  mutate(E=as.numeric(E)+3) %>%
  arrange(E)

ssqBs2=wtSSQ(2)
ssqBs3=wtSSQ(3)

ssqBS<-data.frame(cf=c(2,3),
                  ssq=c(ssq2,ssq3)) %>%
  mutate(compare=ssq/sum(ecDat[,2:ncol(ecDat)]))

ssqBS

```

Variable clutch model:

```{r varClutch}

vClBs<-sapply(ecDat$E,function(x) swiSSQ(2,x))

plot(ecDat$E,vClBs,type="l")

which.min(vClBs)+3

```